---
title: 'EES 3310/5310 Lab #3'
subtitle: "Exercises with the MODTRAN Model"
author: "put your name here"
date: 'Lab: Mon. Jan. 27. Due: Mon. Feb. 3'
fontsize: "12pt"
output:
  html_document: 
    toc: true
  word_document:
    toc: true
  pdf_document:
    toc: true
    includes:
       in_header: ees3310.sty
  github_document:
    toc: true
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)

options(dplyr.summarise.inform = FALSE)

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
data_dir = "_data"
script_dir = "_scripts"

library(pacman)
p_load(xml2, tidyverse, slider)

theme_set(theme_bw(base_size = 15))

source(file.path(script_dir, 'utils.R'), chdir = T)
source(file.path(script_dir, 'format_md.R'), chdir = T)
source(file.path(script_dir, 'modtran.R'), chdir = T)

```
# Chapter 4 Exercises

Fill in R code for the exercises 
(I have put the comment `# TODO` in all of the code chunks where you need to 
do this)
and then fill in the answers where I have marked **Answer:**. 
Be sure to write explanations of your answer and don't just put numbers with 
no text.

## Exercise 4.1: Methane

Methane has a current concentration of 1.7 ppm in the atmosphere and
is doubling at a faster rate than CO~2~.

a) Would an additional 10 ppm of methane in the atmosphere have a larger or 
   smaller impact on the outgoing IR flux than an additional 10 ppm of CO~2~ 
   at current concentrations?

```{r ex_4_1_a, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to explain your reasoning and 
show data, plots, etc. to explain how you came up with your answer.

Suggestion:

* Run MODTRAN in the default configuration (400 ppm CO~2~ and 1.7 ppm methane)
* Run MODTRAN with an extra 10 ppm of CO~2~ and the normal amount of methane.
* Run MODTRAN with the normal amount of CO~2~ and an extra 10 ppm of methane.

What would you look at from the three runs to figure out whether 10 ppm of 
methane or 10 ppm of CO~2~ had the greater effect?


b) Where in the spectrum does methane absorb? What concentration does it take to
  begin to saturate the absorption in this band? Explain what you are looking 
  at to judge when the gas is saturated.
  
**Hints**:  
I recommend setting all the greenhouse gases to zero, and run MODTRAN.
Then run MODTRAN for several values of methane, starting at 1 ppm and 
doubling the concentration until you get to around 128 ppm.
You can do this using a `for` loop, following the examples from the 
lab instructions.

To set all the greenhouse gases to zero, you would call 
`run_modtran(co2_ppm = 0, ch4_ppm = 0, trop_o3_ppb = 0, strat_o3_scale = 0,
h2o_scale = 0, freon_scale = 0)`

The spectrum of methane is complicated so it doesn't saturate all at once.
Use `plot_modtran` to plot the spectrum for each concentration and describe 
what you see and where you think methane begins to saturate and why.

By default, `plot_modtran` gives a plot a title that indicates the CO~2~
concentration. Here, CO~2~ doesn't change and we're interested in the CH~4~
concentration, so you can use the `descr` argument to `plot_modtran` to
give the plots different titles. See the example below.

Remember that if you want to make several plots using a `for` loop, you need to 
assign the result from `plot_modtran` or `ggplot` to a variable and then
use the `plot` or `print` function.
``` r
for (ch4 in ch4_list) {
  mod_data = run_modtran(co2_ppm = 0, ch4_ppm = ch4, trop_o3_ppb = 0,
                         strat_o3_scale = 0, h2o_scale = 0, freon_scale = 0)
  p = plot_modtran(mod_data, descr = str_c(ch4, " ppm methane"))
  plot(p) # you could also say print(p) here.
}
```

```{r ex_4_1_b, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

  
c) Would a doubling of methane have as great an impact on the heat balance as a 
   doubling of CO~2~?

   **Suggestion:** 
    
   * Run MODTRAN in its default configuration (400 ppm CO~2~ and 1.7 ppm methane)
   * Run it again with double the methane concentration.
   * Run it again with the default methane (1.7 ppm) but double the CO~2~ 
     concentration.
   * Compare $I_{\text{out}}$ for the three runs.

```{r ex_4_1_c, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


d) What is the "equivalent CO~2~" of doubling atmospheric methane? That is to 
   say, how many ppm of CO~2~ would lead to the same change in outgoing IR 
   radiation energy flux as doubling methane? What is the ratio of ppm CO~2~ 
   change to ppm methane change?

```{r ex_4_1_d, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


## Exercise 4.2: CO~2~ (Graduate students only)

a) Is the direct effect of increasing CO~2~ on the energy output at the top of
   the atmosphere larger in high latitudes or in the tropics?

   **Hint:**
   Run MODTRAN with the atmosphere set to `tropical`, `midlatitude summer`,
   and `subarctic summer`, and for each atmosphere, at 400 ppm and 800 ppm CO~2~. 
   For each atmosphere, calculate the difference between I~out~ at 400 and 800
   ppm CO~2~ and determine how the effect of doubling CO~2~ varies as you go
   from  tropical latitudes to subarctic ones.


```{r ex_4_2_a, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


b) Set pCO~2~ to an absurdly high value of 10,000 ppm. You will see a spike
   in the CO~2~ absorption band. What temperature is this light coming from? 
   Where in the atmosphere do you think this comes from?

   Now turn on clouds and run the model again. Explain what you see.
   Why are night-time temperatures warmer when there are clouds?

   **Hint:** MODTRAN simulates the upward directed, outgoing longwave radiation 
   as seen by a sensor looking down from some height. 
   For the first part of this exercise, start with the sensor at its
   default altitude of 70 km (you set this with the `altitude_km` argument to
   `run_modtran`), and successively lower it by 10 km at a time until
   you get to 10 km. 
   What happens to the spike as you lower the sensor? What does this say about
   what part of the atmosphere is responsible for the spike in the middle of the 
   CO~2~ absorption at very high values of CO~2~?
   
   For the second part of this exercise, try using "altostratus" clouds.


```{r ex_4_2_b, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.


## Exercise 4.3: Water vapor

Our theory of climate presumes that an increase in the temperature at ground
level will lead to an increase in the outgoing IR energy flux at the top of the
atmosphere.


a) How much extra outgoing IR would you get by raising the temperature of the 
   ground by 5&deg;C? What effect does the ground temperature have on the 
   shape of the outgoing IR spectrum and why?

```{r ex_4_3_a, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

b) More water can evaporate into warm air than into cool air. Change the
   model settings to hold the water vapor at constant relative humidity 
   rather than constant vapor pressure (the default), calculate the change
   in outgoing IR energy flux for a 5&deg;C temperature increase.
   Is it higher or lower? Does water vapor make the Earth more sensitive to
   CO~2~ increases or less sensitive?
   
   **Note:** By default, the MODTRAM model holds water vapor pressure constant,
   but you can set it to hold relative humidity constant instead with the
   option `h2o_fixed = "relative humidity"`, like this:
   `run_modtran(file_name, delta_t = 5, h2o_fixed = "relative humidity")`.

```{r ex_4_3_b, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

c) Now see this effect in another way. 

   * Starting from the default base case, record the total outgoing 
     IR flux. 

   * Now double pCO2. The temperature in the model stays the
     same (that's how the model is written), but the outgoing IR flux
     goes down.

   * Using constant water vapor pressure, adjust the temperature offset
     until you get the original IR flux back again. Record the change in
     temperature
    
   * Now repeat the exercise, but holding the relative humidity fixed
     instead of the water vapor pressure.
    
   * The ratio of the warming when you hold relative humidity fixed
     to the warming when you hold water vapor pressure fixed is the 
     feedback factor for water vapor. What is it?

```{r ex_4_3_c, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

# Chapter 5 Exercise

## Exercise 5.2: Skin Height

a) Run the MODTRAN model using the "tropical" atmosphere, without clouds, 
   and with present-day pCO~2~ (400 ppm). Use the ground temperature reported 
   by the model to calculate
   &epsilon; &sigma; T~ground~^4^,
   the heat flux emitted by the ground.
   Assume &epsilon; = 1, and I have already provided the value of the 
   Stefan-Boltzmann constant &sigma;, as the R variable `sigma_sb`,
   which equals `r format_md(sigma_sb, digits = 3, format = "scientific")`.
   (I defined it in the script "utils.R", which I loaded in the "setup" chunk
   in the RMarkdown document).

   Next, look at the outgoing heat flux at the top of the atmosphere (70 km) 
   reported by the MODTRAN model. Is it greater or less than the heat flux
   that you calculated was emitted by the ground?

```{r ex_5_2_a, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

b) Use the outgoing heat flux at the top of the atmosphere to calcuate the
   skin temperature (use the equation 
   $I_{\text{out}} = \varepsilon \sigma T_{\text{skin}}^4)$).
   What is the skin temperature, and how does it compare to the ground 
   temperature and the temperature at the tropopause, as reported by the 
   MODTRAN model?
   
   Assuming an environmental lapse rate of 6K/km, and using the 
   skin temperature that you calculated above, and the ground temperature
   from the model, what altitude would you expect the skin height to be?

```{r ex_5_2_b, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

c) Double the CO~2~ concentration and run MODTRAN again. Do not adjust the
   ground temperature. Repeat the calculations from (b) of the skin
   temperature and the estimated skin height.

   What is the new skin temperature? What is the new skin height?

```{r ex_5_2_c, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.

d) Put the CO~2~ back to today's value, but add cirrus clouds, using the
   "standard cirrus" value for the clouds. Repeat the calculations from (b) of
   the skin temperature and the skin height.
   
   What is the new skin temperature? What is the new skin height?
   Did the clouds or the doubled CO~2~ have a greater effect on the
   skin height?

```{r ex_5_2_d, include=TRUE, message=FALSE}
# TODO
# Put your R code here
```

**Answer:** _Put your answer here._ Be sure to show your work and
include any data, plots, etc. that you need in order to explain how
you came up with your answer.
